Anomaly detection with kernel density estimation on manifolds

### 2022-05-03 paper/paper.tex
Add S2 example reference

Denote \(B(p, r) \subset T_pM\) as an open ball centered at point \(p\) with raidus \(r\). Then \(B(0_p, r) = exp_p^{-1}(B(p,r))\) is an open neighborhood of \(0_p\) in the tangent space at \(p\), \(T_pM\).
Define \(exp_p\) as the \emph{exponential map} at point \(p\), where \(exp_p\) is a differentiable, bijective map of differentiable inverse (i.e.~\emph{diffeomorphism}) that maps a tangent vector \(u \in B(0_p, r)\) to the endpoint of the geodesic \(\gamma: I \rightarrow M\) satisfying \(\gamma(0)=p, \gamma^{-1}(0)=u\), and \(\gamma(1)=exp_p(u)\).
The exponential map moves point \(p\) from \(p\) at speed \(u\) to an endpoint after covering the length of \(\|u\|\) along the geodesic in one time unit.
The inverse of the exponential map is called the \emph{logarithm map}, \(\log_p(q):= \exp^{-1}_p(q)\), which gives the vector from point \(p\) to \(q\).
Also define the \emph{geodesic ball} centered at \(p\) of radius \(r > 0\) as the image by the exponential map of \(B(0_p, r) \subset T_pM\) with \(r < \textit{inj}_gM\).
Then we could interpolate a geodesic \(\gamma\) between two points \(p\) and \(q\) with the exponential map and the logarithmic map, \(\gamma(t) = \exp_p(t\log_p(q))\), and the geodesic distance is given by \(d_g(p,q) = \|\log_p(q)\|_g\).

\hypertarget{volumn-density-function}{%
\subsection{Volumn density function}\label{volumn-density-function}}

For any \(u \in T_pM\), let \(\gamma(u): t \rightarrow \exp(\frac{tu}{\|u\|})\) and let \(w_i, i=1,\dots,d\) be Jacobi fields along \(\gamma\) such that \(w_i(0)=0\), for all \(i=1,\dots,d\), \(Dw_1/dt(0)=u/\|u\|\) and \(Dw_i/dt(0)\) forms an orthonomal basis of the tangent space \(T_pM\).
Define \(\theta_p(q): T_pM \rightarrow \mathbb{R}\) the volume density function on \(M\) as
\[
\theta_p(u) = \|u\|^{d-1} \det(w_1(\|u\|),\dots,w_d(\|u\|)).
\]
Rewrite \(\theta_p(u) = \theta_p(\log_p(u))\).
The exponential map \(\exp_p: T_pM \rightarrow M\) induced by pullback of a volume form \(\exp_p^*\textit{vol}\) on \(T_pM\), and
\(\theta_p\) is its density with respect to the Lebesgue measure of the Euclidean structure on \(T_pM\), which gives in normal coordinates
\[
d\exp_p^*vol(u) = \theta_p(u)du.
\]

\hypertarget{example}{%
\paragraph{Example}\label{example}}

Manifold: S2

\(\theta \sim U(0, \pi/2)\)

\(X = \cos(\theta)\)

\$\$

>>>>

Denote \(B(p, r) \subset T_pM\) as an open ball centered at point \(p\) with
raidus \(r\). Then \(B(0_p, r) = exp_p^{-1}(B(p,r))\) is an open
neighborhood of \(0_p\) in the tangent space at \(p\), \(T_pM\). Define
\(exp_p\) as the \emph{exponential map} at point \(p\), where \(exp_p\) is a
differentiable, bijective map of differentiable inverse (i.e.
\emph{diffeomorphism}) that maps a tangent vector \(u \in B(0_p, r)\) to the
endpoint of the geodesic \(\gamma: I \rightarrow M\) satisfying
\(\gamma(0)=p, \gamma^{-1}(0)=u\), and \(\gamma(1)=exp_p(u)\). The
exponential map moves point \(p\) from \(p\) at speed \(u\) to an endpoint
after covering the length of \(\|u\|\) along the geodesic in one time
unit. The inverse of the exponential map is called the \emph{logarithm map},
\(\log_p(q):= \exp^{-1}_p(q)\), which gives the vector from point \(p\) to
\(q\). Also define the \emph{geodesic ball} centered at \(p\) of radius \(r > 0\)
as the image by the exponential map of \(B(0_p, r) \subset T_pM\) with
\(r < \textit{inj}_gM\). Then we could interpolate a geodesic \(\gamma\)
between two points \(p\) and \(q\) with the exponential map and the
logarithmic map, \(\gamma(t) = \exp_p(t\log_p(q))\), and the geodesic
distance is given by \(d_g(p,q) = \|\log_p(q)\|_g\).

\hypertarget{volume-density-function}{%
\subsection{Volume density function}\label{volume-density-function}}

For any \(u \in T_pM\), let
\(\gamma(u): t \rightarrow \exp(\frac{tu}{\|u\|})\) and let
\(w_i, i=1,\dots,d\) be Jacobi fields along \(\gamma\) such that \(w_i(0)=0\),
for all \(i=1,\dots,d\), \(Dw_1/dt(0)=u/\|u\|\) and \(Dw_i/dt(0)\) forms an
orthonomal basis of the tangent space \(T_pM\). Define
\(\theta_p(q): T_pM \rightarrow \mathbb{R}\) the volume density function
on \(M\) as \[
\theta_p(u) = \|u\|^{d-1} \det(w_1(\|u\|),\dots,w_d(\|u\|)).
\]
Rewrite \(\theta_p(u) = \theta_p(\log_p(u))\). The exponential
map \(\exp_p: T_pM \rightarrow M\) induced by pullback of a volume form
\(\exp_p^*\textit{Vol}\) on \(T_pM\), and \(\theta_p\) is its density with
respect to the Lebesgue measure of the Euclidean structure on \(T_pM\),
which gives in normal coordinates \[
d\exp_p^*\textit{Vo}l(u) = \theta_p(u)du.
\]

\hypertarget{example-s1}{%
\subsection{\texorpdfstring{Example: \(S^1\)}{Example: S\^{}1}}\label{example-s1}}

Manifold: \(S^1\) with center 0 and radius \(r=1\)

Polar coordinates: \((r, \theta)\)

Uniformly distributed angle: \(\theta \sim U(0, \pi/2)\) with
\(f(\theta) = 2 / \pi\) for \(\theta \in [0, \pi /2]\).

Cartesian coordinates: \((x, y)\) where \(x = \cos(\theta) \in [0, 1]\) and
\(y = \sin(\theta) \in [0, 1]\)

Density of \(\theta\): \(f(\theta) = 2 / \pi\)

Density of \(x\): \(f(x) = \frac{2}{\pi \sqrt{1-x^2}}\)

The CDF of X is given by \[
P(X \leq x) = P(\cos(\theta) \leq x) \\
= P(\theta \geq \arccos(x)) \\
= \int_{\arccos(x)}^{\pi/2} f(\theta) d\theta \\
=  \frac{2}{\pi} (\frac{\pi}{2} - \arccos(x) ) \\
= 1 - \frac{2}{\pi} \arccos(x).
\] Then the density of X is the derivative of the CDF w.r.t. \(x\) \[
f(x) = -\frac{2}{\pi} (-\frac{1}{\sqrt{1 - x^2}}) \\
= \frac{2}{\pi \sqrt{1 - x^2}}.
\] Given \((x, y) \in \mathbb{R}^2\) as sample points from \(S^1\), we could
estimate the density of each sample point \(X_i \in S^1, i = 1, \dots, N\)
and compare it with its true density \(f(x)\).

We use the density estimator in \textcite{Pelletier2005-vu}. For point
\(p \in S^1\), the density at \(p\) is estimated by
\[
f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{X_i}(p)} K(\frac{d_g(p, X_i)}{r}),
\] where \(d=1\), \(h\) is the bandwidth,
\(\theta_{X_i}(p)\) is the volume density function at sample point \(X_i\),
and \(d_g(p, X_i)\) is the geodesic distance between \(p\) and \(X_i\) along
the manifold \(S^1\).

Since the manifold is known (\(S^1\)), we know the geodesic distance is
the arc length between two points on the manifold, which is given by
\(L(\gamma) = r(\beta - \alpha)\) with \(r=1\). Therefore, the geodesic
distance is \[
d_g(p, X_i) = \arccos(p) - \arccos(X_i)
\] As stated in \textcite{Pelletier2005-vu}, in terms of \emph{geodesic normal
coordinates} at \(X_i\), the volume density function \(\theta_{X_i}(p)\)
equals the determinant of the metric \(g\) expressed in these coordinates
at the logarithm map \(\exp_{X_i}^{-1}(p)\), i.e.~\(|g(\exp_p^{-1}(x_i))|\).

{[}When using normal coordinates(orthogonal basis) for tangent vector in
\(T_{X_i}S^1\), the exponential map has \emph{geodesic normal coordinates}.{]}

The Riemannian metric \(g\) for \(S^1\) is given by the inner products on
the tangent spaces of \(S^1\). For each \(X_i \in M\), \[
g_{X_i}: T_{X_i}M \times T_{X_i}M \rightarrow \pmb{R},
\] is smooth in a local neighborhood \(U\) of \(M\).

Then the volume density function \(\theta_{X_i}(p)\) is the inner product
of the tangent vectors produced by the logarithm map from \(X_i\) to \(p\),
which is given by \[
\theta_{X_i}(p) = \sqrt{1 - X_i^2}.
\] Therefore, the estimator for \(p \in S^1\) is given by \[
f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r \sqrt{1 - X_i^2}} K\bigg(\frac{\|\arccos(p) - \arccos(X_i))\|}{r}\bigg),
\] where the bandwidth \(r\) is selected by the plug-in bandwidth selector
implemented by the R package \emph{ks}.

\hypertarget{how-to-derive-volume-density-function-theta_x_ip}{%
\subsubsection{\texorpdfstring{How to derive volume density function \(\theta_{X_i}(p)\)}{How to derive volume density function \textbackslash theta\_\{X\_i\}(p)}}\label{how-to-derive-volume-density-function-theta_x_ip}}

Example of \(S^2\) in \textcite{Henry2009-om}.

Let \(S^2\) be the two-dimensional sphere of radius 1 and \(p \in S^2\). Let
\(v\) and \(w\) be vectors such that \(\|v\| = \|w\| = 1\) and \(\{v, w, p\}\)
is the orthonormal basis of \(\mathbb{R}^3\). Consider the exponential
chart \((U, \Psi)\) induced by the parametrisation
\(\Psi^{-1}: B_\pi(0) \rightarrow S^2 \backslash \{-p\}\) given by \[
\Psi^{-1}(s,t) = \cos(r)p + \sin(r)(sv+tw)r^{-1},
\] if \((s,t) \neq (0,0)\) and \(\Psi^{-1}(0,0)=p\) where
\(r=\sqrt{s^2 + t^2}\). Note that \(r=d_g(p,q)\) if \(q = \exp_p(sv+tw)\).
Then we have \[
A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)}, \\
B(s,t) = \frac{\partial}{\partial \Psi_2} \bigg|_{\Psi^{-1}(s,t)}.
\] Then each component of the riemannian metric could be calculated as
\(g(A(s,t), A(s,t)), g(A(s,t), B(s,t)), g(B(s,t), B(s,t))\). We obtain
that the volume density on the sphere is \[
\theta_p(q) = \bigg| \det g_q\bigg( \frac{\partial}{\partial \Psi_i}\bigg|_q, \frac{\partial}{\partial \Psi_j}\bigg|_q \bigg) \bigg|^{1/2} \\
= \frac{|\sin(d_g(p,q))|}{d_g(p,g)},
\] for \(q \neq p, -p\), and \(\theta_p(p) = 1\).

In \(S^1\), polar coordinates: \(\phi(\exp_p{vt})^{\prime}(0) = v\)

cartesian coordinates:
\(\psi(\exp_p{vt})^{\prime}(0) = x_p \cos{v} - \sqrt{1-x_p^2} \sin{v}\)
\bigg\_t=0 = -\sqrt{1-x_p^2}v\$
